Applying Radial Basis Networks and Markov Chains for on-line detection of concept drift in non-stationary environments

### 2019-10-27 tex/article/article.tex
title and abstract (reviewed)

\begin{abstract}
Most real-world problems experience a phenomenon known as concept drift,
which is a change in data distribution that can affect the system performance.
%
However, the majority of drift detection methods are unfit for non-stationary environments with data streams.
These algorithms or require the correct labeling of data and detect drift through the monitoring of the system performance,
or do not attend the severe response time and resource usage restrictions inherent to data streams.
%
In an attempt to mitigate the aforementioned problem,
this paper proposes a novel proactive method for on-line drift detection,
called RBFChain.
The proposed method relies on Radial Basis Networks implicit clustering property,
besides using Markov Chains to model the transitions and achieve noise tolerance.
%
To assess the proposed method as a viable concept drift detector,
an analysis of sensitivity, accuracy,
and noise tolerance was performed using synthetic datasets,
and results were compared to the most established algorithms in the literature.
%
Furthermore, the algorithm was applied to the real-world problem of eye-tracking.
A relevant problem with impact in different areas of knowledge, since many behavioral experiments use eye-tracking information (fixations and saccades) as a relevant analysis factor.
Performed experiments reveal that RBFChain can classify fixations and saccades in real-time, with high accuracy, noise tolerance, and using limited resources.
\end{abstract}

@@

Still, most concept drift detection methods adopt a reactive approach and require the correct labeling of incoming data.
Generally, this approach is unfit for non-stationary environments with data streams.
In these circumstances,
correct labeling of data can be costly or simply unviable,
and an on-line treatment of data is required since it is a continuous flow.

In an attempt to mitigate the aforementioned problem,
this paper proposes a novel proactive method for on-line drift detection,
called RBFChain.
The proposed method relies on Radial Basis Networks implicit clustering property,
besides using Markov Chains to model the transitions and achieve noise tolerance.

To assess the proposed method as a viable concept drift detector,
an analysis of sensitivity, accuracy,
and noise tolerance was performed using synthetic datasets,
and results were compared to the most established algorithms in the literature.

Furthermore, the algorithm was also tested with the real-world problem of eye-tracking.
This problem is relevant because many behavioral experiments use eye-tracking information (fixations and saccades) as a relevant analysis factor.
Performed experiments reveal that RBFChain can classify fixations and saccades in real-time, with high accuracy, noise tolerance, and using limited resources.

% Eye movements are one of the most natural and repetitive motions of the human being.
% Worldly activities, such as reading a book or watching television, are only possible because of these.
% This regular movement, known as saccadic movement, is required by the retina to obtain a clean image.
% %
% During the analysis of a scene, the eye tends to fixate during a few milliseconds on the most significant areas.
% After this fixation, it moves towards a new area of interest.
% %
% Therefore, it is possible to describe the trajectory of the eye as a sequence of fixation periods interrupted by saccades.
% %
% Fixation periods allow the brain to process information, given that during saccades, the brain does not have time to process the images transmitted by the visual system.
% %
% Hence, the brain assigns semantics to the image only during fixation periods.
% Thus, the study of fixations and saccades is essential for the interpretation and understanding of the eye movements.

% Most commercial eye-tracking systems provide horizontal and vertical coordinates of the point of regard (POR) relative to the screen of a monitor.
% The results presented are composed of sequences of points corresponding to positions of the eyes of an individual.
% The generated eye path consists of eye fixations (with a high density of points) separated by vast spaces where only a few isolated points (saccades) are present.

% Automated identification of fixation and saccades is a necessary aspect in the analysis of visual behavior.
% These classifications serve as underlying knowledge for the metrics used for interpreting eye movements (number of saccades, the average amplitude of saccades, number of fixations, duration of the first fixation, etc).
% %
% Unfortunately, the majority of the algorithms still employ a velocity, acceleration, or dispersion threshold to detect potential saccades \cite{otero:jorge:troncoso:saccades_and_fixations}.
% Some techniques differ by applying additional procedures, such as principal component analysis, to distinguish between smooth search, saccades, and noise \cite{Berg2009}.
% %
% % One exception is the algorithm presented by \cite{Urruty:2007:DEF:1314303.1314308}, which uses dispersion
% % and projection clustering into arbitrary subspace to detect fixations.

% However, advances in the area of behavioral analysis have made measured eye movements more variable and complex, thus making the main methods found in the literature less effective \cite{Anderson:OneAlgorithmToRuleThemAll:2017}.

% This paper proposes a novel algorithm, called RBFChain, which applies Radial Basis Function Networks and Markov Chains to classify scan paths into fixations and saccades in real-time, with high accuracy, noise tolerance, and using limited resources.
% Besides that, experiments conducted in this work show that the true detection rate obtained by RBFChain is comparable to that obtained by state of the art.

% The rest of the paper is organized as follows:
% Section 2 describes the concept drift phenomenon and the main detection techniques;
% Section 3 presents the fixations and saccades detection problem;
% Section 4 describes the RBFChain algorithm and its pseudo-code;
% Section 5 shows the experiment configuration for synthetic datasets  and results obtained;
% Section 6 presents the examination setup with a real-world eye-tracking problem and observed outcomes; and, finally, Section 7 provides conclusions
% and discusses future work.

@@

%
Many relevant real-world problems can be considered as non-stationary environments.
Examples include financial market monitoring, telecom networks, intruder detection, spam filtering, among others \cite{Gama:2014:DAF:2670967.2670971}.

>>>>

\begin{abstract}
Most real-world problems experience a phenomenon known as concept drift,
which is a change in data distribution that can affect the system performance.
%
However, the majority of drift detection methods are unsuited for non-stationary environments with data streams.
These algorithms or require the correct labeling of data - infeasible in these settings,
or do not match the severe response time and resource usage restrictions inherent to data streams.
%
In an attempt to mitigate the aforementioned problem,
this paper proposes a novel proactive method for on-line drift detection,
called RBFChain.
The proposed method relies on Radial Basis Networks implicit clustering property
and uses Markov Chains to model the drifts transitions.
%
To assess the proposed method as a viable concept drift detector,
an analysis of sensitivity, accuracy,
and noise tolerance was performed using synthetic datasets,
and results were compared to the most established algorithms in the literature.
%
Furthermore, the algorithm was applied to the real-world problem of eye-tracking.
A relevant problem with impact in different areas of knowledge, since many behavioral experiments use eye-tracking information (fixations and saccades) as a relevant analysis factor.
Performed experiments reveal that RBFChain can classify fixations and saccades in real-time, with high accuracy, noise tolerance, and using limited resources.
\end{abstract}

@@

Still, most drift detection methods are improper for non-stationary environments with data streams.
These methods or require the correct labeling of data (impracticable in these contexts) and detect drift through the monitoring of the system performance,
or do not meet the severe response time and resource usage restrictions intrinsic to data streams.

This paper proposes a novel proactive method for on-line drift detection,
called RBFChain.
The proposed algorithm is based on Radial Basis Networks implicit clustering property,
and employs Markov Chains to model the transitions and accomplish noise tolerance.

To validate the proposed method as a viable concept drift detector,
an examination of sensitivity, accuracy,
and noise tolerance was performed using synthetic datasets,
and results were compared to the most established algorithms in the literature.

Moreover, the algorithm was also tested with the real-world problem of eye-tracking.
This problem is relevant because many behavioral experiments use eye-tracking information (fixations and saccades) as a relevant analysis factor.
Performed experiments reveal that RBFChain can classify fixations and saccades in real-time, with high accuracy, noise tolerance, and using limited resources.

The rest of the paper is organized as follows:
Section 2 describes the concept drift phenomenon and the main detection techniques;
Section 3 presents the fixations and saccades detection problem;
Section 4 describes the RBFChain algorithm and its pseudo-code;
Section 5 shows the experiment configuration for synthetic datasets  and results obtained;
Section 6 presents the examination setup with a real-world eye-tracking problem and observed outcomes; and, finally, Section 7 provides conclusions
and discusses future work.

@@

Many relevant real-world problems can be considered as non-stationary environments.
Examples include financial market monitoring, telecom networks, intruder detection, spam filtering, among others \cite{Gama:2014:DAF:2670967.2670971}.
